---
title: "Loading and Transforming Global Surface Temperature Data"
author: "Sarah Abdelrahman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This project examines global surface temperature data published by Our World in Data. The dataset provides annual average surface temperatures for countries and regions over time, allowing temperature patterns to be compared across locations and years.

The data and accompanying information are available from Our World in Data: https://ourworldindata.org/grapher/average-annual-surface-temperature

I selected this dataset because climate and temperature data have important geographic and environmental applications. As someone with a background in Geomatics Engineering and GIS who is now studying Data Science, I am interested in how geographic data can be transformed and analyzed using R.

# Loading the Data

The original dataset is loaded directly from Our World in Data. Using a web-based CSV file instead of a file stored on my computer makes the analysis reproducible for other users.

```{r load-data}
temperature_raw <- read.csv(
  "https://ourworldindata.org/grapher/average-annual-surface-temperature.csv"
)

head(temperature_raw)
```

# Data Transformation

For the final data frame, I selected the country or region name, country code, year, and average surface temperature. I renamed the columns to make them easier to understand and to clearly indicate that temperature is measured in degrees Celsius.

```{r transform-data}
temperature_clean <- temperature_raw[, c(
  "Entity",
  "Code",
  "Year",
  "Average.surface.temperature"
)]

names(temperature_clean) <- c(
  "country_or_region",
  "country_code",
  "year",
  "temperature_celsius"
)

head(temperature_clean)

```

# Final Data Frame

The final data frame contains four clearly labeled variables: country or region, country code, year, and average surface temperature in degrees Celsius. The `temperature_celsius` variable is the main outcome of interest, while the other variables provide geographic and temporal information that can be used to compare temperature patterns.

```{r final-data}
str(temperature_clean)
summary(temperature_clean)
```

# Conclusions

This assignment demonstrates how publicly available temperature data can be loaded directly from the web into R and transformed into a data frame with meaningful column names. The resulting data frame is reproducible and provides a simple structure for analyzing changes in surface temperature across countries and years.

To extend this work, I would compare temperature trends among different countries and investigate how average surface temperatures have changed over time. I would also consider creating maps and other visualizations to examine geographic patterns and verify the results using additional climate data sources.
